Destruction of metrizability in generalized inverse limits

### 2015-09-17 articles/destroying-metrizability-in-generalized-inverse-limits/article.tex
update with vargona's latest draft

\documentclass{amsart}

@@

@@

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{game}[theorem]{Game}
\newtheorem{example}[theorem]{Example}
\newtheorem{question}[theorem]{Question}

\newcommand{\<}{\langle}
\renewcommand{\>}{\rangle}

\parskip=0.5em

\title{Destruction of metrizability in generalized inverse limits}

\author{Steven Clontz, Scott Varagona}

\date{\today}

@@

@@

Suppose $X$ and $Y$ are non-empty compact Hausdorff spaces. Then the set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{upper semi-continuous} (u.s.c.) at $x \in X$ if, for every open set $V$ in $Y$ with $f(x) \subseteq V$, there exists an open set $U$ in $X$ with $x \in U$ and $f(u) \subseteq V$ for each $u \in U$. If $f$ is u.s.c. at each $x \in X$, we simply say \emph{$f$ is u.s.c.} If $f: X \rightarrow 2^{Y}$ is a set-valued function, then the \emph{graph of $f$}, denoted $G(f)$, is given by $G(f) = \{\langle x , y \rangle \in X \times Y \ | \ y \in f(x)\}$. (Note that, to help distinguish between ordered pairs and intervals on a linearly ordered space, we will use pointed brackets to denote ordered pairs, e.g., $\langle x , y \rangle$, whereas intervals are given in the traditional way with brackets or parentheses, e.g., $(a,b), [c,d]$, etc.) The set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{surjective} if, for every $y \in Y$, there exists $x \in X$ with $y \in f(x)$. If a u.s.c. function $f: X \rightarrow 2^{Y}$ has the property that $f(x)$ is connected for each $x \in X$, then $f$ is \emph{continuum-valued}, and we denote this by writing  $f: X \rightarrow C(Y).$

@@

Let $\kappa$ be an ordinal. (We follow the convention of set theory in which an ordinal is equal to the set of its predecessors. For more detailed background on ordinals, see, for example, \cite{Kunen}. Assume all ordinals are nonzero.) Suppose that for each $\alpha \in \kappa$, $X_{\alpha}$ is a non-empty compact Hausdorff space and whenever $\alpha \le \beta \in \kappa$, the set-valued function $f_{\alpha \beta} : X_{\beta} \rightarrow 2^{X_{\alpha}}$ is u.s.c. (With this notation, $f_{\alpha \alpha}$ always denotes the identity on $X_{\alpha}$.) Suppose further that, whenever $\alpha \le \beta \le \eta \in \kappa$, we have $f_{\alpha \beta} \circ f_{\beta \eta} = f_{\alpha \eta}$. Then the collection $\textbf{f} = \{ f_{\alpha \beta} \ | \ \alpha \le \beta \in \kappa\}$ of u.s.c. functions is said to be \emph{exact}. In this case, we say $\{X_{\alpha}, f_{\alpha \beta}, \kappa\}$ is an \emph{inverse limit system}; the \emph{inverse limit}, $\varprojlim \{X_{\alpha}, f_{\alpha \beta}, \kappa\}$, of this system is $\{ \textbf{x} \in \prod_{\alpha \in \kappa} X_{\alpha} \ | \ x_{\alpha} \in f_{\alpha \beta}(x_{\beta})$ for all $\alpha \le \beta \in \kappa\}$. (In keeping with tradition, a boldface $\textbf{x}$ always denotes an element of $\prod_{\alpha \in \kappa} X_{\alpha}$; the $\alpha$th coordinate of such an $\textbf{x}$ will be denoted $\textbf{x}(\alpha)$.)  We call the u.s.c. functions $f_{\alpha \beta}$ the \emph{bonding functions} of the inverse limit, and each $X_{\alpha}$ is called a \emph{factor space} of the inverse limit.

@@

We now focus on the case of an inverse limit with a single idempotent, surjective, u.s.c. bonding function $f: [0,1] \rightarrow C([0,1])$. The main goal of this paper is to prove that if such an $f$ is not the identity map, then the inverse limit $\varprojlim \{[0,1], f, \kappa\}$ is metric exactly when the index set $\kappa$ is a countable ordinal.

Let us say a subset $S$ of $[0,1]^2$ satisfies condition $\Gamma$ if, for some $x, y \in [0,1]$ with $x \ne y$, $\<x,x\> \in S$, $\<y,y\> \in S$, and either $\<x,y\>$ or $\<y,x\> \in S$. Let $\Delta = \{\<a,a\> \ | \ a \in [0,1]\}$, and notice that $[0,1]^2 \setminus \Delta$ is the union of two disjoint open sets. Let $\iota$ be the identity map $\iota(x)=\{x\}$.

\begin{lemma}
If $f: [0,1] \rightarrow C([0,1])$ is u.s.c., then for some $x \in [0,1]$, $\<x,x\> \in G(f)$.
\end{lemma}

@@

\begin{proof}
By Lemma 1, for some $x \in [0,1]$, $\<x,x\> \in G(f)$. Suppose by way of contradiction that, for all $y \in [0,1]$ with $y \not= x$, $\<y,y\> \not\in G(f)$.

@@

Now consider $f(1)=[a,b]\subseteq(0,1)$. Then $f(f(1))=f([a,b])=[a,b]$, and in particular, $f(a)\subseteq[a,b]$. However, since $a\in(0,1]$, we also have $f(a)\subseteq(0,a)$, a contradiction.

@@

\textbf{Case 3:}
$x \in (0,1)$. Since $f(0) \subseteq (0,1]$ and the graph of $f$ is connected, $f(z) \subseteq (z,1]$ for each $z \in [0,x)$; thus, $0 \not\in f([0,x))$. Likewise,  since $f(1) \subseteq [0,1)$, we have $f(z) \subseteq [0,z)$ for each $z \in (x,1]$; thus, $1 \not\in f((x,1])$. Since $f$ is surjective,
$1\in f(v)$ for some $v\in[0,x]$ and $0\in f(w)$ for some $w\in[x,1]$. Let us say $f(v) = [k,1]$.

@@

So, $f(1) = [a,b]$ where $[a,b] \subseteq [0,x)$ or $[a,b] \subseteq (x,1)$. If $[a,b] \subseteq (x,1)$, then since $a > x$, $f(a) \subseteq [0,a)$ while $f(a)\subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, contradiction. On the other hand, if $[a,b] \subseteq [0,x)$, then since $b<x$, $f(b) \subseteq (b,1]$ while $f(b) \subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, another contradiction. Since each possible case results in a contradiction, the proof is complete.
\end{proof}

@@

\textbf{Case 1.} There exist $x<y$ in $[0,1]$ such that $\<x,x\>$ and $\<y,y\>$ are elements of $G(f)$ but $\<p,p\>$ is not for all $x<p<y$.

@@

\textbf{Case 2.} There exists $[x,y]\subsetneq[0,1]$ with $x<y$ and $\<p,p\>\in G(f)$ exactly when $p\in[x,y]$. Since $\Delta$ is not a subset of $G(f)$, this means either $0 \not\in [x,y]$ or $1 \not\in [x,y]$. Let us suppose $0\not\in[x,y]$. (The case $1\not\in[x,y]$ is similar.) Choose the minimal $z\in[0,1]$ satisfying $0\in f(z)$. Because $0 \not\in f(0)$, we know $f(0) \subseteq (0,1]$ and thus (because $G(f)$ is connected) $f(p) \subseteq (p,1]$ for $p \in [0,x)$. Thus, $z \in [x,1]$.

@@

\begin{lemma} \label{general lemma}
If $f:[0,1] \to 2^{[0,1]}$ is u.s.c. and idempotent and $G(f)$ satisfies condition $\Gamma$, then $\varprojlim\{[0,1],f,\kappa\}$ contains a copy of $\kappa + 1$.
\end{lemma}

@@

\begin{theorem} \label{main theorem}
If $f:[0,1]\to C([0,1])$ is u.s.c., idempotent, and surjective, but $f\not=\iota$, then $\varprojlim\{[0,1],f,\kappa\}$ contains a copy of $\kappa+1$.
\end{theorem}

@@

\begin{corollary} \label{main corr}
An inverse limit with a single u.s.c., idempotent, surjective bonding function $f: [0,1] \rightarrow C([0,1])$ and index set $\kappa$, an ordinal, is only metric (in fact, Corson compact) in the case that $\kappa$ is countable or the bonding map is trivial.
\end{corollary}

@@

\begin{example} \label{ex1} Let $f: [0,1] \rightarrow C([0,1])$ be given by $f(0) = [0,1]$ and $f(x) = \{1\}$ for each $x \ne 0$. Then $\varprojlim \{ [0,1], f, \kappa \}$ is metric exactly when $\kappa$ is a countable ordinal.
\end{example}

@@

The discussion of this example in \cite{varagona} implies that this inverse limit is homeomorphic to a metric arc whenever $\kappa$ is a limit ordinal satifying $0 < \kappa < \omega_1$, but homeomorphic to the compactified long line when $\kappa = \omega_1$.
\

@@

\begin{example} \label{ex2} Let $f: [0,1] \rightarrow C([0,1])$ be given by $f(x) = \{x\}$ for $0 \le x < 1/2$ and $f(x) = 1/2$ for $x \ge 1/2$.
\end{example}

@@

\bibitem{varagona} Scott Varagona, {\it Generalized Inverse Limits Indexed by Totally Ordered Sets}, preprint, 2015.

>>>>

% This file requires figure.pdf and it should be processed by pdflatex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%           Topology Proceedings Sample Latex Article             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{amsart}
\usepackage[T1]{fontenc}   %%%%%%%Please do not change
\usepackage{graphics}                  %%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please do not change this paragraph. %%%%%%%
\setcounter{page}{1}                   %%%%%%%
\setlength{\textwidth}{4.4in}          %%%%%%%
\setlength{\textheight}{7.0in}         %%%%%%%
\setlength{\evensidemargin}{1in}       %%%%%%%
\setlength{\oddsidemargin}{1in}        %%%%%%%
\setlength{\topmargin}{.8in}           %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please do not change the page size   %%%%%%%
% and do not redefine the baselineskip.%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{question}[theorem]{Question}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This a placeholder for the TOPLOGY PROCEEDINGS logo %%%%%%
\noindent                                             %%%%%%
\begin{picture}(150,36)                               %%%%%%
\put(5,20){\tiny{Submitted to}}                       %%%%%%
\put(5,7){\textbf{Topology Proceedings}}              %%%%%%
\put(0,0){\framebox(140,34){}}                        %%%%%%
\put(2,2){\framebox(136,30){}}                        %%%%%%
\end{picture}                                        %%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{0.5in}

\renewcommand{\bf}{\bfseries}
\renewcommand{\sc}{\scshape}
\newcommand{\<}{\langle}
\renewcommand{\>}{\rangle}
%insert defs/styles
\vspace{0.5in}

\title[Destruction of metrizability in generalized inverse limits]%
{Destruction of metrizability in generalized inverse limits}

%    Information for first author:
\author{Steven Clontz}
\address{Department of Mathematics and Statistics, Fretwell Hall, UNC Charlotte, NC 28223}
%    Current address (if needed):
%\curraddr{}
\email{steven.clontz@gmail.com}
%\thanks{}

%    Information for second author (if needed):
\author{Scott Varagona}
\address{Department of Biology, Chemistry \& Mathematics, University of Montevallo,
Montevallo, AL 35115}
\email{svaragona@montevallo.edu}
%\thanks{Support information for the second author.}

%    General info
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subjclass[2010]{Primary 54F15, 54H20}                                    %
%                                                                                                                           %
%         Please use the current 2010 Mathematics Subject Classification:             %
%         http://www.ams.org/mathscinet/msc/                                                        %
%         http://www.zentralblatt-math.org/msc/en/                                                 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\keywords{Inverse Limits, Inverse Limits with set-valued functions, continuum theory, metrizability, idempotent.}
\thanks {}

@@

Suppose $X$ and $Y$ are non-empty compact Hausdorff spaces. Then the set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{upper semi-continuous} (u.s.c.) at $x \in X$ if, for every open set $V$ in $Y$ with $f(x) \subseteq V$, there exists an open set $U$ in $X$ with $x \in U$ and $f(u) \subseteq V$ for each $u \in U$. If $f$ is u.s.c. at each $x \in X$, we simply say \emph{$f$ is u.s.c.} If $f: X \rightarrow 2^{Y}$ is a set-valued function, then the \emph{graph of $f$}, denoted $G(f)$, is given by $G(f) = \{\langle x , y \rangle \in X \times Y \ | \ y \in f(x)\}$. (Note that, to help distinguish between ordered pairs and intervals on a linearly ordered space, we will use pointed brackets to denote ordered pairs, e.g., $\langle x , y \rangle$, whereas intervals are given in the traditional way with brackets or parentheses, e.g., $(a,b), [c,d]$, etc.) The set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{surjective} if, for every $y \in Y$, there exists $x \in X$ with $y \in f(x)$. If a u.s.c. function $f: X \rightarrow 2^{Y}$ has the property that $f(x)$ is connected for each $x \in X$, then $f$ is \emph{continuum-valued}, and we denote this by writing  $f: X \rightarrow C(Y).$ The \emph{identity on} $X$ is the u.s.c. function $\iota$ given by $\iota(x) = \{x\}$ for all $x \in X$; we may simply say the \emph{identity} (or, more simply, $\iota$) when the space $X$ is understood from context.

@@

Let $\kappa$ be an ordinal. (We follow the convention of set theory in which an ordinal is equal to the set of its predecessors. For more detailed background on ordinals, see, for example, \cite{Kunen}.) Suppose that for each $\alpha \in \kappa$, $X_{\alpha}$ is a non-empty compact Hausdorff space and whenever $\alpha \le \beta \in \kappa$, the set-valued function $f_{\alpha \beta} : X_{\beta} \rightarrow 2^{X_{\alpha}}$ is u.s.c. (With this notation, $f_{\alpha \alpha}$ is always the identity $\iota$ on $X_{\alpha}$.) Suppose further that, whenever $\alpha \le \beta \le \eta \in \kappa$, we have $f_{\alpha \beta} \circ f_{\beta \eta} = f_{\alpha \eta}$. Then the collection $\textbf{f} = \{ f_{\alpha \beta} \ | \ \alpha \le \beta \in \kappa\}$ of u.s.c. functions is said to be \emph{exact}. In this case, we say $\{X_{\alpha}, f_{\alpha \beta}, \kappa\}$ is an \emph{inverse limit system}; the \emph{inverse limit}, $\varprojlim \{X_{\alpha}, f_{\alpha \beta}, \kappa\}$, of this system is $\{ \textbf{x} \in \prod_{\alpha \in \kappa} X_{\alpha} \ | \ x_{\alpha} \in f_{\alpha \beta}(x_{\beta})$ for all $\alpha \le \beta \in \kappa\}$. (In keeping with tradition, a boldface $\textbf{x}$ always denotes an element of $\prod_{\alpha \in \kappa} X_{\alpha}$; the $\alpha$th coordinate of such an $\textbf{x}$ will be denoted $\textbf{x}(\alpha)$.)  We call the u.s.c. functions $f_{\alpha \beta}$ the \emph{bonding functions} of the inverse limit, and each $X_{\alpha}$ is called a \emph{factor space} of the inverse limit.

@@

We now focus on the case of an inverse limit with a single idempotent, surjective, u.s.c. bonding function $f: [0,1] \rightarrow C([0,1])$. The main goal of this paper is to prove that if such an $f$ is not the identity and $\kappa$ is a nonzero ordinal, then the inverse limit $\varprojlim \{[0,1], f, \kappa\}$ is metric exactly when the index set $\kappa$ is countable.

Let us say a subset $S$ of $[0,1]^2$ satisfies condition $\Gamma$ if, for some $x, y \in [0,1]$ with $x \ne y$, $\<x,x\> \in S$, $\<y,y\> \in S$, and either $\<x,y\>$ or $\<y,x\> \in S$. Let $\Delta = \{\<a,a\> \ | \ a \in [0,1]\}$, and notice that $[0,1]^2 \setminus \Delta$ is the union of two disjoint open sets.

\begin{lemma} \label{first lemma}
If $f: [0,1] \rightarrow C([0,1])$ is u.s.c., then for some $x \in [0,1]$, $\<x,x\> \in G(f)$.
\end{lemma}

@@

\begin{proof}
By Lemma \ref{first lemma}, for some $x \in [0,1]$, $\<x,x\> \in G(f)$. Suppose by way of contradiction that, for all $y \in [0,1]$ with $y \not= x$, $\<y,y\> \not\in G(f)$.

@@

Now consider $f(1)=[a,b]\subseteq(0,1)$, where $a \le b$. (If $a = b$, then $[a,b] = \{a\}$ and the following argument will still apply.) Then $f(f(1))=f([a,b])=[a,b]$, and in particular, $f(a)\subseteq[a,b]$. However, since $a\in(0,1]$, we also have $f(a)\subseteq(0,a)$, a contradiction.

@@

\textbf{Case 3:}
$x \in (0,1)$. Since $f(0) \subseteq (0,1]$ and the graph of $f$ is connected, $f(z) \subseteq (z,1]$ for each $z \in [0,x)$; thus, $0 \not\in f([0,x))$. Likewise,  since $f(1) \subseteq [0,1)$, we have $f(z) \subseteq [0,z)$ for each $z \in (x,1]$; thus, $1 \not\in f((x,1])$. Since $f$ is surjective, $1\in f(v)$ for some $v\in[0,x]$ and $0\in f(w)$ for some $w\in[x,1]$. If $f(v) = \{1\}$, then $f(f(v)) = f(\{1\}) = \{1\}$, which is a contradiction since $\<1,1\> \not\in G(f)$. So, let us say $f(v) = [k,1]$ for some $k < 1$.

@@

So, $f(1) = [a,b]$ for some $a \le b$, where $[a,b] \subseteq [0,x)$ or $[a,b] \subseteq (x,1)$. If $[a,b] \subseteq (x,1)$, then since $a > x$, $f(a) \subseteq [0,a)$ while $f(a)\subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, contradiction. On the other hand, if $[a,b] \subseteq [0,x)$, then since $b<x$, $f(b) \subseteq (b,1]$ while $f(b) \subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, another contradiction. Since each possible case results in a contradiction, the proof is complete.
\end{proof}

@@

\textbf{Case 1.} $G(f) \cap \Delta$ is not connected. Then there exist $x<y$ in $[0,1]$ such that $\<x,x\>$ and $\<y,y\>$ are elements of $G(f)$ but for all $p$ with $x < p < y$, $\< p, p \> \not\in G(f)$.

@@

\textbf{Case 2.} $G(f) \cap \Delta$ is connected. Then there exists $[x,y]\subsetneq[0,1]$ with $x<y$ and $\<p,p\>\in G(f)$ exactly when $p\in[x,y]$. Since $\Delta$ is not a subset of $G(f)$, this means either $0 \not\in [x,y]$ or $1 \not\in [x,y]$. Let us suppose $0\not\in[x,y]$. (The case $1\not\in[x,y]$ is similar.) Choose the minimal $z\in[0,1]$ satisfying $0\in f(z)$. Because $0 \not\in f(0)$, we know $f(0) \subseteq (0,1]$ and thus (because $G(f)$ is connected) $f(p) \subseteq (p,1]$ for $p \in [0,x)$. Thus, $z \in [x,1]$.

@@

\begin{lemma} \label{general lemma}
Let $\kappa$ be a nonzero ordinal. If $f:[0,1] \to 2^{[0,1]}$ is u.s.c. and idempotent and $G(f)$ satisfies condition $\Gamma$, then $\varprojlim\{[0,1],f,\kappa\}$ contains a copy of $\kappa + 1$.
\end{lemma}

@@

\begin{theorem} \label{main theorem}
Let $\kappa$ be a nonzero ordinal. If $f:[0,1]\to C([0,1])$ is u.s.c., idempotent, and surjective, but $f\not=\iota$, then $\varprojlim\{[0,1],f,\kappa\}$ contains a copy of $\kappa+1$.
\end{theorem}

@@

\begin{corollary} \label{main corr}
An inverse limit with a single u.s.c., idempotent, surjective bonding function $f: [0,1] \rightarrow C([0,1])$ and index set $\kappa$, a nonzero ordinal, is only metric (in fact, Corson compact) in the case that $\kappa$ is countable or the bonding map is trivial.
\end{corollary}

@@

\begin{example} \label{ex1} Let $\kappa$ be a nonzero ordinal, and let $f: [0,1] \rightarrow C([0,1])$ be given by $f(0) = [0,1]$ and $f(x) = \{1\}$ for each $x \ne 0$. Then $\varprojlim \{ [0,1], f, \kappa \}$ is metric exactly when $\kappa$ is countable.
\end{example}

@@

The discussion of this example in \cite{varagona} implies that this inverse limit is homeomorphic to $Y = (\kappa \times [0,1)) \cup \{\< \kappa, 0 \>\}$ (with the lexicographic order topology) whenever $\kappa$ is a nonzero limit ordinal. If $0 < \kappa < \omega_1$, $Y$ is a metric arc. However, when $\kappa = \omega_1$, $Y$ is the compactified long line, a non-metric space.
\

@@

\begin{example} \label{ex2} Let $f: [0,1] \rightarrow C([0,1])$ be given by $f(x) = \{x\}$ for $0 \le x < 1/2$ and $f(x) = \{1/2\}$ for $x \ge 1/2$.
\end{example}

@@

\bibitem{varagona} Scott Varagona, {\it Generalized Inverse Limits Indexed by Totally Ordered Sets}, preprint (submitted to Topology Proceedings), 2015.
